Progress Report on One Simulated Dataset

### 2023-10-23 Docs/ProgressReportSim.Rmd
example code for one iteration of simulation

- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly.
- I would like to use time points on the [0, 1] scale as evaluation points when creating basis functions.
- Also, take number of knots and order from the FPCA model in step 3 (knots = 35, cubic spline basis). 
- Do the same kind of projection to mean functions as well

@@

# project mean functions 
lm_mod <- lm(fpca_mod$mu ~ B-1)
mu_new <- Bnew %*% coef(lm_mod)
```

@@

```{r}
plot(mid_t, fpca_mod$mu, xlab="bin", ylab="Mean", pch=20, cex = 0.5)
lines(t, mu_new, col="blue")
```

- The projection seems to work very well!

@@

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction. And out-of-sample scores are not estimated by FPCA but Laplace approximation.
- I will use the projected eigenfunctions from above. 
- I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation

@@

@@

\begin{aligned}
\phi_k(t) & = \sqrt{n_{bin}}\hat{\phi}_k(t)  \\
\lambda_k & = \hat{\lambda}_k/n_{bin}

\end{aligned}

@@

- Re-scaled eigenvalues
```{r}
new_lambda/n_bin
```

@@

plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")

@@

```{r, results='hide'}
# approximation process
# container
pred_list <- list()
# score <- array(NA, dim=c(length(test_id), K, 4))
# dimensions: subject, eigenfunction, maximum observation time

@@

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_new[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(new_lambda), f0=mu_new[1:max_rid, ])

@@

    # prediction
    eta_pred_out <- mu_new+efunctions_new%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

@@

<!-- ### Compared to biased eigenvalues -->

```{r, results='hide'}
## With eigenvalues from FPCA
# container
pred_list2 <- list()

@@

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_new[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(fpca_mod$evalues[1:K]), f0=mu_new[1:max_rid, ])

    # prediction
    eta_pred_out <- mu_new+efunctions_new%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  pred_list2[[i]] <- df_i
}
```

```{r}
df_pred2 <- bind_rows(pred_list2)

df_pred2$pred0.2[df_pred2$t<=0.2] <- NA
df_pred2$pred0.4[df_pred2$t<=0.4] <- NA
df_pred2$pred0.6[df_pred2$t<=0.6] <- NA
df_pred2$pred0.8[df_pred2$t<=0.8] <- NA
# overview of predicted track
p2<-df_pred2 %>%
  filter(id %in% rand_test_id) %>%
  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "Unscaled, biased eigenvalues")

```{r, results='hide'}
## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
# container
pred_list3 <- list()
efunctions_scaled <- efunctions_new*sqrt(n_bin)
evalues_scaled <- new_lambda/n_bin

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_scaled[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(evalues_scaled), f0=mu_new[1:max_rid, ])

    # prediction
    eta_pred_out <- mu_new+efunctions_scaled%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

@@

```{r, fig.height=5, fig.width=15}
ggarrange(p1, p2, p3, nrow = 1, common.legend = T)
```

@@

```{r}
# calcualte ISE
err2 <- df_pred2 %>%
  mutate(err1 = (pred0.2-eta_i)^2,
         err2 = (pred0.4-eta_i)^2,
         err3 = (pred0.6-eta_i)^2,
         err4 = (pred0.8-eta_i)^2) %>%
  select(id, t, starts_with("err"))

err2$window = cut(err2$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
# table(err1$window)
err2 <- split(err2, f = err2$window)

tb2 <- lapply(err2, function(x){
  x %>%  group_by(id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>%
    summarise_at(vars(err1, err2, err3, err4), mean)
}) %>% bind_rows(.id = "Window")
```

@@

ise <- full_join(tb1, tb2, by = "Window") %>% full_join(tb3, by = "Window")
colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 3))
ise %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 12))
```

@@

```{r}
# calcualte AUC
auc_df2 <- df_pred2 %>%
  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
  select(Y, starts_with("pred"), window)

auc_mat2 <- matrix(NA, length(pred_window), length(obs_track))

for(j in seq_along(obs_track)){
  for(i in (j+1):5){
    this_Y <- auc_df2$Y[auc_df2$window==pred_window[i]]
    this_pred <- auc_df2[auc_df2$window==pred_window[i],
                         paste0("pred", obs_track[j])]
    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
    auc_mat2[i ,j] <- this_perf@y.values[[1]]
  }
}

@@

auc <- cbind(auc_mat, auc_mat2, auc_mat3)
rownames(auc) <- unique(auc_df1$window)
colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 3))
auc %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 12))
```

@@

### BAM

- The bam model below couldn't do out-of-sample prediction
- It also take a long time to fit

```{r,class.source='show', eval=FALSE}
# try to fit a flexible GLMM
t1 <- Sys.time()
fit_glmm_sp <- bam(Y ~ s(t, bs="cr", k=10)+ti(id, t, bs=c("re", "cr"), k = c(5, 5), mc = c(TRUE, FALSE)), 
                   family = binomial, data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
```

- I should construct basis functions and fit model with GLMMadaptvie::mixed_model instead

### GLMMadaptive + basis functions

@@

>>>>

- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly.
- I would like to use time points on the [0, 1] scale as evaluation points when creating basis functions.
- Also, take number of knots and order from the FPCA model in step 3 (knots = 35, cubic spline basis). 
- We don't project the mean function here. Instead, we will use the re-evaluated mean function from the next part (bam model used for debias)

@@

# project mean functions 
# lm_mod <- lm(fpca_mod$mu ~ B-1)
# mu_new <- Bnew %*% coef(lm_mod)
```

@@

@@

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates) and the mean function. They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction. And out-of-sample scores are not estimated by FPCA but Laplace approximation.
- I will use the projected eigenfunctions from above. 
- I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation

@@

We will need the mean function from this model, derived below:

This is pretty much a linear functions. Is that supposed to happen?

```{r, class.source = "fold-show"}
new_mu <- predict(debias_glmm, type = "terms")[, 1]
new_mu <- unique(new_mu)
plot(unique(train_df$t), new_mu, type = "l")
```

@@

\[\begin{aligned}
\phi_k(t) & = \sqrt{n_{bin}}\hat{\phi}_k(t)  \\
\lambda_k & = \hat{\lambda}_k/n_{bin}
\end{aligned}\]

@@

- Re-scaled eigenvalues

```{r}
scaled_lambda <- new_lambda/n_bin
```

@@

```{r, fig.height=8, fig.width=8}
efunctions_scaled <- efunctions_new*sqrt(n_bin)

par(mfrow=c(2,2))

plot(t, efunctions_scaled[, 1], type="l", xlab="bin", ylab="PC1")
plot(t, efunctions_scaled[, 2], type="l", xlab="bin", ylab="PC2")
plot(t, efunctions_scaled[, 3], type="l", xlab="bin", ylab="PC3")
plot(t, efunctions_scaled[, 4], type="l", xlab="bin", ylab="PC4")

@@

```{r pred_debias_unscale, results='hide'}
# approximation process
# container
pred_list <- list()
# score <- array(NA, dim=c(length(test_id), K, 4))
# dimensions: subject, eigenfunction, maximum observation time

@@

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_new[1:max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(new_lambda), f0=new_mu[1:max_rid])

@@

    # prediction
    eta_pred_out <- new_mu+efunctions_new%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

@@

```{r pred_debias_scale, results='hide'}
## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
# container
pred_list3 <- list()

@@

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_scaled[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- new_mu+efunctions_scaled%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

@@

```{r, fig.height=5, fig.width=10}
ggarrange(p1, p3, nrow = 1, common.legend = T)
```

@@

@@

ise <- full_join(tb1, tb3, by = "Window")
colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 2))
ise %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 8))
```

@@

@@

```{r}
# calcualte AUC
options(knitr.kable.NA = "")

auc <- cbind(auc_mat, auc_mat3)
rownames(auc) <- unique(auc_df1$window)
colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 2))
auc %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 8))
```

## Investigate a bit more on the difference in prediction

- The scaled and unscaled parameters seem to make a difference in prediction, which they shouldn't

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2, 2))
plot(df_pred$pred0.2, df_pred3$pred0.2, pch=20, cex = 0.5, 
     xlab="unscaled", ylab="scaled", main = "Up to 0.2")
plot(df_pred$pred0.4, df_pred3$pred0.4, pch=20, cex = 0.5,
     xlab="unscaled", ylab="scaled", main = "Up to 0.4")
plot(df_pred$pred0.6, df_pred3$pred0.6, pch=20, cex = 0.5,
     xlab="unscaled", ylab="scaled", main = "Up to 0.6")
plot(df_pred$pred0.8, df_pred3$pred0.8, pch=20, cex = 0.5,
     xlab="unscaled", ylab="scaled", main = "Up to 0.8")
```

@@

@@

Also tried increasing number of basis functions, which ended up with "computationally singular". In fact, this method does not seem very numerically stable.
